Our Members

### 2017-03-13 pnsac/newsletter/article_EditorsNotes.tex
Incorporated changes from first draft review.

This past year started off as a repeat of the difficult year the
Association experienced in 2014. Consequently there was little news
and less enthusiasm for publishing the Chronicle on any regular
basis. However, as a result of changes to the management of the museum
organization there is a renewed sense of hope that we will be able to
move forward in 2016, as expressed in the Notes from the
President. One of those changes is the appointment of Chris Kitzan as
the new director general of CASM. You will find a message from Chris
in this edition. Chris and his boss Alex Benay, the CEO of the museum
corporation, are enthusiastic supporters of our organization. We are also
commencing a new feature highlighting the members of our Association
and starting off with Bruce Gemmill, our Project Manager. In addition,
we are adding a "Miscellany" feature which lists items of interest to
our members. If any of our readers wish to contribute an article, have
suggestions for future articles, have suggestions for the "Miscellany"
feature, or have any comments about the Chronicle please let me know.

>>>>

As the President has noted the Association has, through its volunteers,
expended thousands of hours on the North Star's restoration since 2004. In the
last couple of years, however, various developments at CASM have resulted in a
diminution in the amount of restoration work we have been able to do. The result
has been a decline in the number of volunteers working on the aircraft, and a
corresponding decline in our membership. This in turn has led to an overall
decline in the Association's activities, including the publication of this
newsletter. CASM has made some major changes to its personnel, including those
directly responsible for the restoration of the North Star. The Association is
already experiencing a positive change in its relationship with the Museum which
can only be to the good. We look forward to a renewed partnership with CASM to
our mutual benefit.

### 2017-03-13 pnsac/newsletter/article_Feature1.tex
Incorporated changes from first draft review.

\title{Volunteers Working on North Star Restoration}
%\subtitle{Part 4}
\author{Tom MacGregor, Legion Magazine}

@@

\begin{quotation}
	You have to be careful opening up the panels covering the 
	engines, you never know if a bird or some animal might have made
	its home there.
\end{quotation}

That’s one lesson a group of dedicated volunteers has learned in its
10 years of painstaking work restoring a Royal Canadian Air Force
North Star aircraft at the Canada Aviation and Space Museum in Ottawa.

@@

The aircraft had sat neglected outside the museum for
years. “Essentially, the RCAF flew this one into the Rockcliffe Airport
when it was decommissioned. It sat in a old until Robert Holmgren said
someone should do something about it,” said project manager Bruce
Gemmill.

The airplane had been stripped of its non-essential gear and turned
over to what was then the National Aeronautical Collection, housed in
hangars built during the Second World War at the former RCAF station
at Rockcliffe in Ottawa. Time and weather took its toll. Birds and
animals made nests where they could get inside the panelling.

“We approached the museum about restoring the aircraft. At first, they
didn’t know how to take us,” said Gemmill. “The museum had its own
professional conservation staff but it had never worked with a team of
volunteers.”

Eventually the museum agreed to work with the volunteers who formed
the Project North Star Association of Canada with the late Robert
Holmgren as president. “We became the model for other volunteer groups
now working with the museum,” said the association’s current
president, Richard Lodge.

Now the aircraft gets towed into a hangar where it is available to the
volunteers between September and May before it is towed outside for
the summer.

The North Star was built by Canadair in the late 1940s and 1950s. It
was the RCAF version of the DC-4 civilian aircraft except that it used
four Rolls-Royce Merlin engines. Canadair built 71 of the aircraft.

The RCAF assigned the rst North Stars to 412 Squadron where they
transported VIPs and were used in various transfigurations for
reliable, long-range transport services.

During the Korean War the North Star was used by No. 426 (Thunderbird)
Sqdn. to ferry sup- plies across the Pacific Ocean to Japan (Operation
Hawk: The Korean Airlift, July/August). They would y 599 round trips
over the Pacific and deliver seven million pounds of cargo and 13,000
personnel on return trips. All this was achieved without a fatal
crash.

Most of the air force’s North Stars were declared surplus in the
1970s.

In the 10 years that Project North Star has been in operation, the
group has restored two of the four Merlin engines with a third one
nearly completed.

Each engine had to be taken o the aircraft, brought into the museum’s
shops where they were taken apart. Each item was tagged, catalogued 
and usually photographed. “We take a lot of photographs. That’s
how we know how they go back together,” said Lodge.

Engine project leader Garry Dupont said, “The first engine took us about
four and a half years to restore. We’re almost finished the third engine
which as taken us about two years. So we are getting faster.” The
propellers were restored separately by Hope Aero Propeller and
Components Inc. in Mississauga, Ont. All restoration work has to be up
to the professional standards set by the museum.

The group plans to restore the aircraft to its original look, not to
have it y again.

Lodge said that they have some RCAF manuals but a lot of time they do
not help when it comes to putting 50-year-old parts together today.

The aircraft itself, 17515, has seen countless hours of work put into
re- storing the cockpit with all its switches and navigation
equipment and a small galley for the crew of seven.

Still to be done is work in the cargo section and painting the
exterior of the aircraft.

“We know this aircraft ew during the Korean War. In fact, we have 
pictures showing it configured with litters to bring the wounded back to
Canada,” said Gimmell. “We don’t want to put in the litters but we
have people working on creating the rope seats that were used.”

The plan is to paint the aircraft to its post-1965 markings, since it
was painted with the Maple Leaf ag on it.

Still the search goes on to nd new members for the association. “The
important thing is to nd new people to whom we can pass on this 
knowledge and skills we are developing,” said Lodge.

>>>>

\title{Reunion of Giants}
%\subtitle{Part 4}
\author{Bruce Grant}

@@

The "Reunion of Giants" video came to the Canadian Aviation and Space Museum's
theatre on May 14th, courtesy of the Canadian Warplane Heritage Museum in
Hamilton. CWHM sent along the pilot Captain Leon Evans (retired Air Canada), and
Dave Rohrer, CEO, as well as Morgan Elliott the videographer. Also present were
two World War Two decorated Lancaster pilots Don McKechnie and Bill Button. With
almost a full house of 200 it was a very successful Project North Star public
event.

The outstanding feature of the story is the audacity of CWHM to fly Canada's
only airworthy Lancaster to England for a reunion with the only other flying
Lancaster in the world. The Lanc is the flagship of the CWHM, its biggest
attraction and an irreplaceable historic artifact. In a typical year it flies
only about fifty hours, attending aviation events in Canada and every move is
calculated some years in advance.

In England, the arrival over the North Atlantic route of our Lanc was embraced
as a tribute to the quality of the machine and the courage of its crew. It
reinforced the link with Canada which provided so much support to Great Britain
during the war.

While on a final approach in England, the supercharger on engine \#3 went to
pieces and the plane landed with a trail of smoke. A replacement engine would
have to be found. The replacement that turned up was a Rolls Royce Merlin that
would have to be substituted for the Packard Merlin in our Canadian-built
Lancaster. This presented a challenge in sorting out the elements of QEC with
incompatible Whitworth and SAE fasteners. The new engine would be a "bitsa"
(bitsa this and bitsa that). They did it in about three days and got the plane
back in the air.

On the return flight, stronger than anticipated headwinds raised the risk of
running out of gas over the Davis Strait. They diverted to Narsarsuaq in
Greenland.

Following the showing of the film there was a spirited question and answer
session with CWHM's representatives moderately by Mike Pearson (retired Air
Canada pilot of the Gimli glider fame). There is no doubt that all those in
attendance enjoyed themselves thoroughly and gained some understanding of what
it meant for those involved in this project.

At Project North Star, we have some appreciation idea of the scale of CWHM's
accomplishment. As our members know we are in the twelfth year of our project to
restore the last surviving North Star. But, we are restoring it only as a museum
artifact, not an airworthy machine. The video is available for sale at
www.warplane.com and the trailer can be viewed ath www.suddenlyseemore.com .

%\begin{quotation}
%	\textit{You have to be careful opening up the panels covering the 
%	engines, you never know if a bird or some animal might have made
%	its home there.}
%\end{quotation}

@@

%\textit{Reprinted from the March/April 2014 edition of "Legion Magazine," with the kind permission of Jennifer Morse, General Manager of Canvet Publications, and Tom MacGregor.}

### 2017-03-13 pnsac/newsletter/article_MemberInterview.tex
Incorporated changes from first draft review.

\title{Our Members}
\author{Interview with Bruce Gemmill}

@@

\textit{Bruce Gemmill has been involved with Project North Star since 2005 and
	has exceeded 9,000 volunteer hours. Bruce is presently the Project
	Manager, as well as a member of the board of directors.}

>>>>

\title{Our Members}
\author{Interview with Jm Riddoch}

@@

\begin{quotation}
	\textit{Jim Riddoch is one of the earliest and longest serving members of the
Association. He has made a significant contribution to the operations of the
Association including working on the aircraft as well as being an officer and
member of the board of directors. Although officially retired from formal
positions in the organization, he can still be counted on to help out when
needed.}
\end{quotation}

1.\textit{What is your background in aviation?}

I started with English Electric Aviation Company in 1956 as an
apprentice technician eventually being assigned to the Mechanical
Engineering Department as a fuel systems technician. I remained there
for 5 years until I was 28 years old and married with a daughter. After
immigrating to Canada in March 1966 I joined DeHavilland Spar Division
in Malton and then moved to Montreal with Jarry Hydraulics as a
development engineer. I remained there for 2 years working chiefly on
development proposals for the wing sweep actuators for F111 aircraft
and Boeing's proposed supersonic Concorde.

In 1968 I joined Air Canada as a junior engineer in the Mechanical
Systems Department and assigned as a landing gear technologist until I
achieved Professional Engineer status in 1974. I then moved around
several other departments in engineering and maintenance until I was
promoted to Superintendent of DC9 maintenance. In that capacity I was
involved in the accident investigation into the loss of one aircraft at
Cincinnati.

I eventually returned to engineering as Director of Interior Systems and
Equipment to supervise extensive fleet modifications and new aircraft
acquisitions. I remained in that position until I retired in 1990.
Following a short stint with First Air I was asked to join a newly formed
Canadian Aviation Maintenance Council to standardize aircraft
maintenance trades basic training. I was hired as the Accreditation
Manager to approve various training establishments including schools,
companies and military programs, I eventually also took on the job of
Registration Manager to accept qualified trade technicians.

2. \textit{How long have you been involved with Project North Star and how and why did you get involved?}

Following my retirement from the Canadian Aviation Maintenance
Council in 2003 I was approached by Robert Holmgren to participate in
a voluntary program to help restore an the Canadair North Star aircraft
parked outside at the Canada Aviation Museum. Along with Tim
Timmins and a few others we formed a steering committee to approach
the Museum about forming a voluntary work force to assist the museum
staff to restore the aircraft. At first we met with opposition from some
staff members who felt their jobs were being compromised and possible
staff reductions. Slowly we were approved by the Museum staff with
limited supervision and controlled access and work direction. I was
assigned by the Project North Star Association as Chief Engineer of the
project although all work was controlled and supervised by the
Museum's staff manager.

%\begin{figure}[htbp]
%   \vspace{2em}
%   \centering
%   %name of the graphic, without the path AND in EPS format:
%   \includegraphics[scale=0.5]{fwd_cockpit_complete.eps}
%   %caption of the figure 
%   \caption*{\small \em Forward cockpit -- complete.}
%   %label of the figure, which has to correspond to \ref{}:
%   \label{fig:fwd_cockpit_complete}
%\end{figure}

3. \textit{What has been the history of your involvement to date?}

I worked on the Project for 10 years starting in 2003 both as a “hands
on” volunteer as well as a member of the board of directors and officer
of the Association. Generally I put in at least two days per week and
sometimes more.

%\begin{figure}[htbp]
%	\vspace{2em}
%	\centering
%	%name of the graphic, without the path AND in EPS format:
%	\includegraphics[scale=0.9]{resources/PNS2008-10-26_23-31-42.png}
%	%caption of the figure 
%	\caption*{\small \em Propeller assembly.}
%	%label of the figure, which has to correspond to \ref{}:
%	\label{fig:propeller}
%\end{figure}

%\begin{figure}[htbp]
%	\vspace{2em}
%	\centering
%	%name of the graphic, without the path AND in EPS format:
%	\includegraphics[scale=1.3]{resources/NorthStarCockpitDec2011.png}
%	%caption of the figure 
%	\caption*{\small \em Forward cockpit -- complete.}
%	%label of the figure, which has to correspond to \ref{}:
%	\label{fig:fwd_cockpit_complete}
%\end{figure}

4. \textit{What has been the highlight of your involvement?}

The chief highlights have been the high level of workmanship of the
volunteers and displaying the aircraft to the public and interested
persons.

%\begin{figure}[htbp]
%   \vspace{2em}
%   \centering
%   %name of the graphic, without the path AND in EPS format:
%   \includegraphics[scale=0.75]{resources/NavigatorRack.png}
%   %caption of the figure 
%   \caption*{\small \em Navigator rack.}
%   %label of the figure, which has to correspond to \ref{}:
%   \label{fig:nav-radio-rack}
%\end{figure}

5. \textit{What has been the most challenging part of your involvement?}

Keeping the Museum staff and management on side with our goals
and objectives and recruiting enough new volunteers to keep the
project going.
